---
title: "Mini-Project 2: Forecasting US Energy Consumption"
author: Kimberly Williams
format: 
  html: 
    embed-resources: true
editor: visual
---

You are asked to predict the monthly US residential energy consumption for the next 2 years for the 'Short-Term Energy Outlook' using data from the US Department of Energy.

```{r}
#| include: false
# load packages
library(tidyverse)
library(astsa)
```

## IMPORT

The US Department of Energy collects an impressive amount of data to support the energy industries. For this assignment use the monthly total energy consumed by the US residential sector. The webpage

<https://www.eia.gov/totalenergy/data/browser/index.php?tbl=T02.02>

contains monthly web tables, but an easier approach to getting the data into R uses the result of the 'Download' button to read the source data from the source CSV file.

```{r}
#| warning: false
energy <- read_csv("https://www.eia.gov/totalenergy/data/browser/csv.php?tbl=T02.02",
# names in first row, but the names aren't meaningful
                   col_names = c("MSN","Time","Res_Energy","label1","label2","label3"),
                   skip = 1) |>
  mutate(Res_Energy = parse_number(Res_Energy),
         Time = ym(Time))
```

## TIDY

```{r}
glimpse(energy)
```

Prompt: Why is `energy` tidy?

The dataset energy is tidy because: 1) Observations are in rows, 2) Variables are in columns, 3) there is only one number in every chunk so each individual observation listed only has one number per variable.

*Note:* Your answer requires a response to each of the 3 tidy data rules.

## TRANSFORM

Code should

-   filter to the data in rows where `MSN` is `TERCBUS` (Total Energy Consumed by the Residential Sector in trillion Btu),

-   filter to current data (after 1991),

-   remove the annual total rows (`Time` is missing / `NA`).

-   only include the columns `Time` and `Res_Energy` columns, and

```{r}
energy <- energy |>
# filter to TERCBUS Total Energy Consumed by the Residential Sector
  filter(MSN == "TERCBUS") |>
# filter to "since the 90's"
  filter(Time >= ymd("1992-01-01")) |>
# remove annual total rows (Time is NA)
  na.omit() |>
# keep just the Time and Res_Energy columns
  select(Time, Res_Energy)

# take a look
tail(energy)
```

## VISUALIZE

Plot the time series for US residential energy consumption (in trillion Btu).

**Note:** You will need to set `eval: true` when you have created the `energy` dataset in the TRANSFORM section.

```{r}
ggplot(energy, aes(x = Time, y = Res_Energy)) +
  geom_line() +
  labs(
    title = "US Residential Energy Consumption Over Time",
    x = "Time", 
    y = "Residential Energy Usage (in Trillian BTU)"
    )
```

## MODEL

Fit an ARIMA$(1, 1, 1) \times (1, 1, 1)_{12}$ model and report the parameter estimates and standard errors.

**Note:** You will need to set `eval: true` when you have created the `energy` dataset in the TRANSFORM section.

```{r}
energy_out <- sarima(energy$Res_Energy, 1,1,1, 1,1,1, 12)

# table of estimates and se
energy_out$ttable
```

Forecast the US residential energy consumption for the next 2 years. Report a table of predictions and 95% prediction intervals.

**Note:** You will need to set `eval: true` when you have created the `energy` dataset in the TRANSFORM section.

```{r}
#  "next two years" 
energy_forecast <- data.frame(Time = max(energy$Time) + months(1 : 24))
energy_yhat <- sarima.for(energy$Res_Energy , n.ahead = 24, 1,1,1, 1,1,1, 12)

# table of forecasts and uncertainty
energy_forecast$pred <- energy_yhat$pred
energy_forecast$L <- energy_yhat$pred - 2 * energy_yhat$se
energy_forecast$U <- energy_yhat$pred + 2 * energy_yhat$se

energy_forecast
```

## COMMUNICATE

**Prompt:** Does it appear there is a monthly pattern? If so, explain the possible business/economic 'cause' for this effect.

There does seem to be a monthly pattern with the winter months taking significantly more energy than the summer months and fall and spring being less than both winter and summer. The likely explanation for the pattern in energy consumption is temperature regulation. Every building I've been to with heating and cooling units typically set their temp to a balmy 72ºF. Winter takes a lot of energy because of the amount required to heat a building starting from 0º to 40ºF up to 72ºF. Summer takes second most because of the amount required to cool a building starting from 80º to 100ºF down to 72ºF. Whereas Fall and Spring are typically around 50º to 80ºF so it takes less energy to heat or cool it to people's preferred temp. Winter is also probably more extreme because of the need to increase electrical lighting usage. Most other things that require electricity are pretty consistent from season to season, so heating and cooling makes the most sense to me.

```{r}
tail(energy, n=18)
```
