---
title: "HW 2: Palmer Island Penguins"
author: Kimberly Williams
format: 
  html: 
    embed-resources: true
editor: visual
---

```{r}
#| label: load-packages
#| include: false
library(tidyverse)
library(palmerpenguins)
```

## 1. Palmer Penguins

Measurements for penguin species, island in Palmer Archipelago, size (flipper length, body mass, bill dimensions), and sex.

**Import:**

```{r}
penguins
```

**Visualize:**

Make a plot showing differences between penguins and their bills.

*Hint*: Look at the "Coding Out Loud" lecture for the graphic that answers this question.

*Note*: Replace the `"___"` and uncomment (remove the `#`) to see the effect

```{r}
ggplot(penguins,
       aes(x = bill_depth_mm,
           y = bill_length_mm,
           color = species)) +
  geom_point() +
  labs(
    title = "Bill depth and length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Bill Depth (mm)", y = "Bill Length (mm)",
    color = "Species",
    caption = "Source: Palmer Station LTER / palmerpenguins package"
    ) +
    scale_color_viridis_d() 
```

## 2. Chinstrap Penguins

Chinstrap penguins are one of the species in the Palmer Archipelago.

**Import:**

```{r}
chinstrap <- penguins |>
  filter(species == "Chinstrap") |>
  na.omit()
chinstrap
```

**Visualize flipper length:**

*Hint*: Since penguin flipper length is univariate data, to answer this question you will want to try a histogram, density plot, boxplot and your answer will be the one of them that you think best communicates the shape, center, variation, and anything unusual about the distribution.

*Note*: Change `eval` to `true` and replace the `"___"`

```{r}
#| eval: true
ggplot(chinstrap, aes(x = flipper_length_mm)) +
  geom_density(adjust = 0.8, fill = "darkorchid", alpha = 0.5) +
  labs(
    x = "Flipper Length",
    y = "Density",
    title = "Chin Strap Flipper Length"
  )
chinstrap <- penguins |>
  filter(species == "Chinstrap") |>
  na.omit()
chinstrap
median(chinstrap$flipper_length_mm)
IQR(chinstrap$flipper_length_mm)
```

**Communicate:**

Prompt: What should someone know about the Chinstrap penguin flipper length distribution? (shape, center, spread, anything unusual)

-   Shape: Normal, nearly symmetric, with a tiny skew toward larger values. Uni-model
-   Center: 196mm
-   Spread: Most of the data is between 180 and 210 with an IQR of 10mm
-   No unusual observations

**Visualize distribution of year (data collection over time):**

*Note*: Change `eval` to `true` and replace the `"___"`

```{r}
#| eval: true
ggplot(chinstrap, aes(x = year)) +
  geom_histogram(bins = 3, color = "black", fill = "darkred", alpha = 0.7)+
  labs( 
    x = "Year", 
    y = "Count", 
    title = "Chinstraps Over Time"
  ) 
```

**Communicate:**

Prompt: What should someone know about how the Chinstrap penguin data was collected over time (different years)?

The number of penguins collected was different from year to year. The first and third years are around 25, but the second year was less than 20.

## Stretch Exercise:

Is the Chinstrap penguin flipper length distribution approximately normal?

Yes, it is approximately normal.

```{r}
ggplot(data = chinstrap, aes(x = flipper_length_mm))+
  geom_density(fill = "lightpink", alpha = 0.5) +
  stat_function(fun = dnorm, 
                args = list(mean = mean(chinstrap$flipper_length_mm),
                            sd = sd(chinstrap$flipper_length_mm)),
                color = "purple") +
  labs( 
  x = "Flipper length (mm)",
  title = "Chinstrap Flipper length Distribution",
  subtitle = "Penguins on Palmer Islands",
  caption = "Source: Palmer Station LTER / palmerpenguins package"
  )
```
