Recap of last lecture
  Computing New Columns
  tidyverse (mutate)
  Base R ($)

Analysis: Understanding Hotel Prices
Hotel Pricing Demand
Goal is to predict the price a consumer would purchase a hotel room
  If offered price is too high, the hotel will lose revenue because the room goes unused.
  If the offered price is too low, the consumer will pay for the room because they got a good deal but the hotel will have lost potential revenue because they could have charged a higher price and still got the booking.

Open: Wrangling4YourTurn.qmd

Tidy Tuesday: they upload a new, already tidy, data set every day.

Import:
```{r}
library(tidyverse)

hotel <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-11/hotels.csv')
```


Tidy:
```{r}
slice_sample(hotel, n = 10)
# or A different view of the provided dataset
glimpse(hotel)
```

Understand:
pay less by staying longer?
weekend stay costs more at City Hotel?

Transform:
Dataset is tall and wide. What are we focused on?
  Room Cost
  Focus on the City Hotel Property
  Focus on the Bookings not cancelled
  Do we have length of stay? How do we compute it?
  Do we know which bookings include a weekend? How do we compute it?
```{r}
cityhotel <- hotel |>
  filter(hotel == "City Hotel") |>
  filter(is_canceled == 0) |>
  mutate(stay = stays_in_weekend_nights + stays_in_week_nights,
         weekend = ifelse(stays_in_weekend_nights > 1, "Weekend", "NotWeekend")) |>
  select(adr, weekend, stay)
```

Dataset for analysis
```{r}
slice_sample(cityhotel, n = 15)
```

Visualize:
Pay Less by Staying Longer?
```{r}
ggplot(cityhotel, aes(x = stay, y = adr)) +
  geom_point() +
  geom_smooth() +
  labs(title = "City Hotel")
```
```{r}
cityhotel |>
  filter(stay < 20) |>
  filter(adr > 10) |>
  ggplot(aes(x = stay, y = adr)) +
    geom_jitter(size = 3, color = "tan", alpha = 0.2) +
    geom_smooth(se = FALSE) +
    scale_y_log10() +  
      labs(title = "City Hotel",
           y = "Ave Daily Rate ($)",
           x = "Length of Stay (days)")
```

Weekend Stay Costs More?
```{r}
#This doesn't work
ggplot(cityhotel, aes(x = adr, fill = weekend)) +
  geom_density(alpha = 0.5) +
  labs(title = "City Hotel")
```
```{r}
#This does
ggplot(cityhotel, aes(x = weekend, y = adr)) +
  geom_boxplot() +
  labs(title = "City Hotel",
       y = "Ave Daily Rate ($)",
       x = NULL)
```


Transform:
  Add some lines to the previous creation of cityhotel
  Log transformation can’t have adr zero
  additional columns
```{r}
cityhotel <- hotel |>
  filter(hotel == "City Hotel") |>
  filter(is_canceled == 0) |>
  filter(adr > 10) |>
  mutate(stay = stays_in_weekend_nights + stays_in_week_nights,
         weekend = ifelse(stays_in_weekend_nights > 1, "Weekend", "NotWeekend")) |>
  select(adr, weekend, stay, lead_time, is_repeated_guest, 
         required_car_parking_spaces, total_of_special_requests)

slice_sample(cityhotel, n = 15)
```



