## R Basics 2

Recap of last lecture
  Atomic Vectors
  Lists
  
Outline
  Data Frames
  Modifying Values
  Loops and Vectorized Code / Functional Programming
  Logical Subsetting
  Missing Information

## Data Frames
One Factor Experiment
```{r}
library(tidyverse)
burn <- read_csv("
Order, Color, Time
Run1, Yellow, 270
Run2, Black, 177
Run3, Red, 121
Run4, Red, 200
Run5, Yellow, 259
Run6, Red, 125
Run7, Black, 118
Run8, Yellow, 233
Run9, Black, 154
")
```
Looks like a matrix, but it’s a list!
```{r}
dim(burn)
```
```{r}
typeof(burn)
```
```{r}
attributes(burn)
```


## Selecting Values from Data Frames
```{r}
burn
```
```{r}
burn[4, 1]
```
```{r}
burn[4, c(2, 3)]
```
```{r}
burn[4, ]
```
Notice indexing begins at 1

```{r}
burn[, 3]
```
```{r}
burn$Color
```
```{r}
typeof(burn$Color)
```

```{r}
burn |>
  filter(Color == "Yellow")
```
```{r}
burn[burn$Color == "Yellow", ]
```
```{r}
burn[burn$Color == "Yellow"] #THIS DOESN'T WORK BECAUSE THERE IS NO COMMA, look for the correct one up above
```



## Modifying Values
Changing Values in Place
```{r}
y <- c(0, 0, 0, 0, 0, 0)
y
```
Select the first value of y
```{r}
y[1]
```
Modify it
```{r}
y[1] <- 1000
y
```
Replace multiple values
```{r}
y[c(1, 3, 5)] <- c(1, 11, 111)
y
```
```{r}
y[c(2, 4)] <- -4
y
```
Values that do not exist yet
```{r}
y[7 : 9] <- c(7, 8, 9)
y
```
```{r}
z <- NULL
z
```
```{r}
z <- c(z, 5 : 2)
z
```




## Loops and Vectorized Code
Working with a dataset: Many ways to do the same thing
Compute Stock Return

Disney Stock Price
```{r}
DIS <- read_csv("https://query1.finance.yahoo.com/v7/finance/download/DIS?period1=-252374400&period2=9999999999&interval=1d&events=history&includeAdjustedClose=true") |>
  select(Date, Close)

tail(DIS, 12)
```
tidyverse
```{r}
DIS1 <- DIS |>
  mutate(logreturn = log(Close / lag(Close)) )
tail(DIS1)
```

```{r}
head(DIS, 12)
```
for loop
```{r}
DIS2 <- DIS
for(i in 1 : dim(DIS)[1]){
  if(i == 1){
    DIS2$logreturn[1] <- NA
  } else{
    DIS2$logreturn[i] <- log(DIS$Close[i] / DIS$Close[i - 1])
  }
}
head(DIS2)
```

Compute Stock Return
Which is faster? tidyverse vs for loop
tidyverse
```{r}
# Start the clock
ptm <- proc.time()

DIS1 <- DIS |>
  mutate(logreturn = log(Close / lag(Close)) )

# Stop the clock
proc.time() - ptm
```
for loop
```{r}
# Start the clock
ptm <- proc.time()

DIS2 <- DIS
for(i in 1 : dim(DIS)[1]){
  if(i == 1){
    DIS2$logreturn[1] <- NA
  } else{
    DIS2$logreturn[i] <- log(DIS$Close[i] / DIS$Close[i - 1])
  }
}

# Stop the clock
proc.time() - ptm
```




## Loops and Functional Programming
Working with a dataset: Many ways to do the same thing
Compute Annual Maximum

Coal Creek
```{r}
bigdata <- read_table("https://nwis.waterservices.usgs.gov/nwis/iv/?sites=10242000&parameterCd=00065&startDT=2000-01-01T00:00:00.000-06:00&endDT=2026-12-31T23:59:59.999-06:00&siteStatus=all&format=rdb",
                      col_names = c("label1", "label2", "Date", "Time", "tz", "Height", "label3"),
                      skip = 32) |>
  mutate(Date = ymd_hms(paste(Date, Time), tz = "America/Denver")) |>
  select(Date, Height)

tail(bigdata)
```

Conceptualize Computation
```{}
      2007    2008           2020    2021
                     . . .                     . . .
     Height  Height         Height  Height
      4.77    5.51           4.28    4.46
      4.79    4.83           4.38    4.77
      4.74   13.03           4.51    4.47 
      4.76    5.08           4.28    5.43
      4.87    4.91           4.93    4.94
       .       .              .       .
       .       .              .       .
       .       .              .       .

Max: 14.2    13.9    . . .   6.94    9.68      . . .
```


Which is faster?
tidyverse
```{r}
# Start the clock
ptm <- proc.time()

peak1 <- bigdata |>
  mutate(Year = year(Date)) |>
  group_by(Year) |>
  summarize(
    Max_Height = max(Height)
  )

# Stop the clock
proc.time() - ptm
```
for loop
```{SQL}
FOR each value of YEAR

  FILTER dataset to observations of YEAR
  
  COMPUTE max(Height) for filtered dataset
  
END
```

```{r}
# Start the clock
ptm <- proc.time()

my_years <- unique(year(bigdata$Date))
peak2 <- rep(0, length(my_years))
for(i in 1 : length(my_years)){
  this_year <- filter(bigdata, year(Date) == my_years[i])
  peak2[i] <- max(this_year$Height)
}

# Stop the clock
proc.time() - ptm
```



## Logical Subsetting
```{r}
1 > 2
```
```{r}
1 > (0 : 2)
```
```{r}
(1 : 3) == (3 : 1)
```
%in%
```{r}
1 %in% c(3, 4, 5)
```
```{r}
c(1, 2) %in% c(3, 4, 5)
```
```{r}
c(1, 2, 3, 4) %in% c(3, 4, 5)
```

```{r}
x <- 3
x > 2 & < 9
```

```{r}
x <- 3
2 < x & x < 9
```
```{r}
a <- c(1, 2, 3)
b <- c(1, 2, 3)
c <- c(1, 2, 4)
a == b
```
```{r}
b == c
```
```{r}
a == b & b == c
```



## Missing Information
```{r}
y <- c(NA, 1 : 20)
y
```
 [1] NA  1  2  3  4  5  6  7  8  9 10 11 12 13 14 15 16 17 18 19 20

na.rm
```{r}
mean(y)
```
[1] NA
```{r}
mean(y, na.rm = TRUE)
```
[1] 10.5

is.na
```{r}
c(1, 2, 3, NA) == NA
```
[1] NA NA NA NA
```{r}
z <- c(1, 2, 3, NA)
is.na(z)
```
[1] FALSE FALSE FALSE  TRUE
```{r}
z <- c(NA, z)
z

[1] NA  1  2  3 NA