Simulation Studies

Recap of last lecture
  Simulating Data from a Probability Model

Outline
  Simulation Study Examples:
    Normal Goodness of Fit Tests
    Mini-Project: Simulation Study

Assessing Normality
  Experience looking at histograms / boxplots / density plots
  Using GOF tests
    Shapiro-Wilk Test
    K-S test can be more generally applied
    
Applying Normal Goodness of Fit Test
Define Ho: Normal
  Normal(mu, sigma)
    f(y) = 1/(sigma*sqrt(2π)) * exp[(y-mu)^2 / -2sigma^2]
  Features: continuous, symmetric, 68% within mu ± sigma, 99.7% within mu ± 3sigma
Define Ha: Not Normal
  At least one Feature that make the distribution non-normal
  Test statistic looks for difference in features (skewed, discrete, multimodal, heavy/long tails, etc)
  Note: Won’t know what the non-normal distribution is!
  Poisson(lambda = 5)

Shapiro-Wilk test
  Definition of P(Type 1 Error)
  Probability of Rejecting Ho when Ho is true
  Probability of declaring data as non-normal given data is from a normal
  Simulation to Approximate P(Type 1 Error)
Pseudo Code
```{r}
Assign value for n

Create SW_Type1, an empty vector of length Nsim

for(i in 1 : Nsim){
  
  generate n observations from N(0, 1)
  
  compute Shapiro-Wilk Test

  save p-value in SW_Type1[i]
  
}

Compute the proportion of SW p-values < 0.05
```

Shapiro-Wilk test
Definition of Power
  Probability of Rejecting Ho when Ha is true
  Probability of declaring data as non-normal when it is from a non-normal distribution
Note: Need to choose a specific non-normal distribution and be clear on all details
More precise to say “Probability of declaring data as non-normal when it is Poisson(5) with n = 15”

Simulation to Approximate Power
Pseudo Code for Ha: Poisson(5)
```{r}
Assign value for n

Create SW_Power, an empty vector of length Nsim

for(i in 1 : Nsim){
  
  generate n observations from Poisson(5)
  
  compute Shapiro-Wilk Test

  save p-value in SW_Power[i]
  
}

Compute the proportion of SW p-values < 0.05
```
Note: why is N(0,1) WLOG but Poisson(5) is not? because S-W test adjusts for location-scale (mean - variance), but it does not adjust for shape

Code for approx P(Type I error)
```{r}
n <- 15

Nsim <- 10^5

pvals <- rep(0, Nsim)

for(i in 1 : Nsim){
  y <- rnorm(n)
  out <- shapiro.test(y)
  
  pvals[i] <- out$p.value
  
}

mean(pvals < 0.05)
```

Code for approx Power
```{r}
n <- 15

Nsim <- 10^5

pvals <- rep(0, Nsim)

for(i in 1 : Nsim){
  y <- rpois(n, lambda = 5)
  out <- shapiro.test(y)
  
  pvals[i] <- out$p.value
  
}

mean(pvals < 0.05)
```

You may think about other power calculations that are interesting
Sample Size: n = 10, 15, 30, 50, 100
Non-Normal Distributions: Poisson(lambda): lambda = 20, 10, 5 (as lambda increases the Poisson can be approximated by a normal)
Don’t those look like factors / levels from a designed experiment?

Approx P(Type 1 Error)
```{r}
SWtest_Type1Error <- function(n, Nsim = 10^5){
  pvals <- rep(0, Nsim)

  for(i in 1 : Nsim){
    y <- rnorm(n)
    out <- shapiro.test(y)
  
    pvals[i] <- out$p.value
  }
  mean(pvals < 0.05)
}
```
```{r}
SWtest_Type1Error(n = 10)
```

Approx Power (Poisson)
```{r}
SWtest_Power <- function(n, lambda, Nsim = 10^5){
  pvals <- rep(0, Nsim)

  for(i in 1 : Nsim){
    y <- rpois(n, lambda)
    out <- shapiro.test(y)
  
    pvals[i] <- out$p.value
  }
  mean(pvals < 0.05)
}
```
```{r}
SWtest_Power(n = 10, lambda = 5)
```

Power of Shapiro-Wilk test
  Results:
  Prompt: How well do Normal Goodness of Fit tests work?
  Prompt: How dramatic does the non-normal distribution need to be for the test to detect normality?
  Prompt: How large a sample is needed?

Lecture Review
  Simulation study Examples
    Normal Goodness of Fit Tests
Appendix:
  Extracting results (returned values) from R functions


Appendix
Extracting results (returned values) from R functions
```{r}
y <- rpois(15, lambda = 5)
  
out <- shapiro.test(y)

out
```
```{r}
names(out)
```
```{r}
out$p.value
```
```{r}
shapiro.test
```
