## Working With Multiple Datasets
Recap of last lecture
  Information in Multiple Datasets (Relational Databases)
  Keys
  Joins
  
Outline
  Information in Multiple Datasets
  Stacking Datasets
  Case Study: UCLA Cancer Patients
  
Goal
  We have multiple datasets
  want to bring them together
  
BYU Football TV Audience
  1984 National Champions
  Why does BYU play football?
      A way for Alumn to connect with current students
      Missionary opportunity
      BYU brand
  Each Year BYU plays multiple games
  Each year’s data is in a separate dataset

Libraries
```{r}
library(tidyverse)
```


BYU 2022
```{r}
BYU2022 <- read_csv("
Opponent, TVaud, Where, Network, When
South Florida, NA, Away, ESPNU, Saturday
Baylor, 2370, Home, ESPN, Saturday
Oregon, 2580, Away, FOX, Saturday
Wyoming, 609, Home, ESPN2, Saturday
Utah State, 1260, Home, ESPN, Thursday
Notre Dame, 2530, Neutral, NBC, Saturday
Arkansas, 1060, Home, ESPN, Saturday
Liberty, NA, Away, ESPNU, Saturday
East Carolina, 823, Home, ESPN2, Friday
Boise State, 281, Away, FS2, Saturday
Utah Tech, NA, Home, BYUtv, Saturday
Stanford, 453, Away, FS1, Saturday
SMU, 2040, Neutral, ABC, Saturday
")
BYU2022
```

BYU 2021
```{r}
BYU2021 <- read_csv("
Opponent, TVaud, Where, Network
Arizona, 933, Neutral, ESPN
Utah, 1503, Home, ESPN
Arizona State, 1498, Home, ESPN
South Florida, 491, Home, ESPN2
Utah State, NA, Away, CBSSN
Boise State, 2244, Home, ABC
Baylor, 1440, Away, ESPN
Washington State, 466, Away, FS1
Virginia, 678, Home, ESPN2
Idaho State, NA, Home, BYUtv
Georgia Southern, NA, Away, ESPN+
USC, 1458, Away, ESPN
UAB, 3221, Neutral, ABC
")
BYU2021
```

BYU 2020
```{r}
BYU2020 <- read_csv("
Opponent, TVaud, Where, Network
Navy, 1150, Away, ESPN
Troy, 795, Home, ESPN
Louisiana Tech, 505, Home, ESPN2
UTSA, 317, Home, ESPN2
Houston, 1160, Away, ESPN
Texas State, 716, Home, ESPN
Western Kentucky, 788, Home, ESPN
Boise State, 680, Away, FS1
North Alabama, NA, Home, BYUtv
Coastal Carolina, 1210, Away, ESPNU
San Diego State, 606, Home, ESPN2
UCF, 1550, Neutral, ESPN
")
BYU2020
```

bind_rows()
```{r}
BYU <- bind_rows(BYU2022, BYU2021, BYU2020)
slice_sample(BYU, n = 4)
```

BaseR rbind()
```{r}
BYU1 <- rbind(BYU2021, BYU2020)
slice_sample(BYU1, n = 4)
```

Base R rbind() --> error caused when more columns in data set than another data set
```{r}
rbind(BYU2022, BYU2021, BYU2020)
```

bind_rows() Details
```{r}
BYU <- bind_rows(BYU2022, BYU2021, BYU2020)
slice_sample(BYU, n = 4)
```
What happened to Year ?
It was never a column (it was in the name)

```{r}
BYU <- bind_rows(BYU2022, BYU2021, BYU2020, .id = "Year")
print(BYU, n = 38)
```

A few ways to fix this
```{r}
BYU <- bind_rows(BYU2022, BYU2021, BYU2020, .id = "Year") |>
  mutate(Year = case_when(Year == 1 ~ 2022,
                          Year == 2 ~ 2021,
                          Year == 3 ~ 2020))
print(BYU, n = 38)
```
```{r}
BYU <- bind_rows("2022" = BYU2022,
                 "2021" = BYU2021,
                 "2020" = BYU2020, .id = "Year") |>
  mutate(Year = parse_number(Year))
print(BYU, n = 38)
```


## Visualize
Boxplot
```{r}
BYU |>
  mutate(Where = factor(Where, levels = c("Home", "Away", "Neutral"))) |>
  ggplot(aes(x = Where, y = TVaud)) +
    geom_boxplot(color="black", fill="#0062BB") +
    labs(
      x = NULL,
      y = "Viewers (1,000s)",
      title = "BYU Football TV Audience"
    ) +
    theme_minimal()
```

Line
```{r}
BYU |>
  group_by(Year) |>
  summarize(Total = sum(TVaud, na.rm = TRUE) / 10^3) |>
  ggplot(aes(x = Year, y = Total)) +
    geom_line(linewidth = 3, color="#0062BB") +
    geom_point(shape = 21, size = 5, stroke = 5, color="#0062BB", fill = "white") +
    scale_x_continuous(breaks = 2020 : 2022) +  
    labs(
      x = NULL,
      y = "Season Total Viewers (million)",
      title = "BYU Football TV Audience"
    ) +
    theme_minimal()
```








## Case Study: UCLA 
## Cancer Patient Data
Case Study: UCLA Cancer Patient Data
Data Available:
  Patient Data from UC Hospitals
  Patient Data from Cedars-Sinai Hospital
  Doctor Data
Desired Dataset:
  First Stack Patient Data from Different Hospitals
  Join Doctor Data to Patient Data

# Import
UC Patients
```{r}
uc <- read_csv("https://stats.idre.ucla.edu/stat/data/rdm/patient_pt1_dm.csv")

slice_sample(uc, n = 10)
```

Cedars-Sinai Patients
```{r}
cs <- read_csv("https://stats.idre.ucla.edu/stat/data/rdm/patient_pt2_dm.csv")

slice_sample(cs, n = 10)
```

Doctors
```{r}
doc <- read_csv("https://stats.idre.ucla.edu/stat/data/rdm/doctor_dm.csv")

doc
```

# Transform (to make Tidy)
Patient Data joined with Doctor Data
```{r}
cancer <- uc |>
  bind_rows(cs) |>
  inner_join(doc)

slice_sample(cancer, n = 10)
```

Choosing a Doctor is Hard
  Want Old (seen and practiced so much) or Young (current training) ?
  Response variable: experience
  Cancer characteristics: tumorsize, ntumors, cancerstage
  Patient characteristics: age, sex, mobility
  Doctor characteristics: medicaid, lawsuits
  
#Visualize
Tumor size
```{r}
ggplot(cancer, aes(x = tumorsize, y = experience)) +
  geom_point() +
  geom_smooth()
```

Number of tumors
```{r}
ggplot(cancer, aes(x = ntumors, y = experience)) +
  geom_point() +
  geom_smooth()
```

Cancer stage
```{r}
ggplot(cancer, aes(x = cancerstage, y = experience)) +
  geom_boxplot()
```

Age (WRONG)
```{r}
ggplot(cancer, aes(x = age, y = experience)) +
  geom_point() +
  geom_smooth()
```
Age (no outliers)
```{r}
cancer |>
  filter(age < 100) |>
  ggplot(aes(x = age, y = experience)) +
    geom_point() +
    geom_smooth()
```

Sex (WRONG)
```{r}
ggplot(cancer, aes(x = sex, y = experience)) +
  geom_boxplot()
```
Sex (no outliers)
```{r}
cancer |>
  filter(sex %in% c("female", "male")) |>
  ggplot(aes(x = sex, y = experience)) +
    geom_boxplot()
```

Mobility
```{r}
ggplot(cancer, aes(x = mobility, y = experience)) +
  geom_point() +
  geom_smooth()
```

Medicaid
```{r}
ggplot(cancer, aes(x = medicaid, y = experience)) +
  geom_point() +
  geom_smooth()
```

lawsuits
```{r}
ggplot(cancer, aes(x = factor(lawsuits), y = experience)) +
  geom_boxplot()
```



# Transform
```{r}
cancer1 <- cancer |>
  filter(age < 100) |>
  filter(sex %in% c("female", "male")) |>
  mutate(cancerstage = factor(cancerstage)) |>
  mutate(cancerstage = relevel(cancerstage, ref = "IV")) |>
  mutate(nolawsuits = ifelse(lawsuits == 0, TRUE, FALSE)) |>
  select(experience,
         tumorsize, ntumors, cancerstage, age, sex, mobility, 
          medicaid, nolawsuits)
cancer1
```


#Model
```{r}
out <- lm(experience ~ ., data = cancer1)

summary(out)
```


Case Study: UCLA Cancer Patient Data
Communicate
  - Higher mobility scores have more experienced doctors.
  - Doctors who haven’t had lawsuits haven’t practiced long (pvalue < 0.0001). 
      One interpretation is that lawsuits are part of a doctor’s career.
  - Experienced doctors have more Medicaid patients in their practice.
  - More experienced doctors have patients with larger and more tumors, 
      earlier stage cancers, and older patients (pvalues near 0.05).

Lecture Review
  Information in Multiple Datasets
  Stacking Datasets
  Case Study
